---
title: "Using FAIR Theory for Causal Inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using FAIR Theory for Causal Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


<!-- In this example, we illustrate how to use DAGs for causal inference, including the detection of a violation of the initial model and subsequent adaptation of the DAG. We could use that to illustrate updating FAIR theory: -->

<!-- https://currentprotocols.onlinelibrary.wiley.com/doi/full/10.1002/cpz1.45 -->

<!-- We can find more examples of causal inference with DAGs in these tutorials: -->

<!-- https://www.r-bloggers.com/2019/08/causal-inference-with-dags-in-r/ -->

<!-- https://www.r-bloggers.com/2018/08/applications-of-dags-in-causal-inference/ -->


## Introduction

Directed acyclic graphs (DAGs) are a powerful tool for expressing and testing causal assumptions.
They allow researchers to identify potential confounders or colliders,
and guide decisions about which variables to control for (or not) in statistical analyses [@cinelliCrashCourseGood2022].

DAGs can be implemented as FAIR theories, or can be derived from FAIR theories.

In this vignette, we'll illustrate how to use DAGs for causal inference in `R`,
inspired by an example from [@asendorpfBiasDueControlling2012].
In his commentary, Asendorpf discusses the common practice of controlling for variables that may actually be colliders, leading to biased estimates.
He illustrates this principle with data from @motti-stefanidiAdaptationWellbeingAdolescent2012.
The `theorytools` package contains a synthetic dataset based on the summary statistics reported in Motti-Stefanidi and colleagues, `motti-stefanidi2012`.



## Background: Collider Bias in Research on Immigrant Adaptation

Asendorpf (2012) reflects on the study by Motti-Stefanidi and colleagues, which investigated immigrant students’ adaptation in Greece.
He argues that **Immigrant Status** and **Self-Efficacy** both influence **Greek Language Skills**. When researchers study the association between **Immigrant Status** and **Self-Efficacy** (as a proxy for adaptation to the host country),
the results are biased if control for **Greek Language Skills**.
This occurs because **Greek Language Skills** are a collider, or common effect, of the relationship between **Immigrant Status** and **Self-Efficacy**, not an independent cause of self-efficacy.

In this tutorial, we use this example to:

- [ ] Illustrate how one might specify a DAG as a FAIR theory (i.e., imagine you are the theoretician)
- [ ] Access that FAIR theory in the data analysis environment (i.e., imagine you are a user of someone else's theory)
- [ ] Use that FAIR theory to select control variables
- [ ] Perform causal inference

<!-- Below, we’ll translate this example into a simple DAG, simulate some data, and demonstrate how controlling for the collider can change the observed relationship between Immigrant Status and Self-Efficacy. -->

## Step 1: Install and Load Required Packages

We'll use two popular packages for DAGs:

- **dagitty** (for creating and analyzing DAGs)
- **tidySEM** (for visualizing DAGs)

```{r install-packages, eval=FALSE}
install.packages("dagitty")
# install.packages("ggdag")
install.packages("tidySEM")
```

```{r load-libraries}
library(dagitty)
library(tidySEM)
library(dplyr)
library(ggplot2)
```

## Step 2: Specify the DAG

We'll encode the causal relationships as described in Asendorpf’s example:

- **Immigrant Status** -> **Greek Language Skills**
- **Self-Efficacy** -> **Greek Language Skills**
- Potential direct link between **Immigrant Status** and **Self-Efficacy**.

```{r dag-spec}
immigration_dag <- dagitty("dag {
  ImmigrantStatus -> GreekLanguageSkills
  SelfEfficacy -> GreekLanguageSkills
  ImmigrantStatus -> SelfEfficacy
}")

graph_sem(immigration_dag)

dagitty::localTests(immigration_dag, sim_data)
```
The DAG shows that **Greek Language Skills** is a collider because it has two parents: **Immigrant Status** and **Self-Efficacy**.

## Step 4: Simulate Some Data

Below, we’ll simulate a small dataset where:
1. **ImmigrantStatus** is binary (0 = non-immigrant, 1 = immigrant)
2. **SelfEfficacy** is correlated with ImmigrantStatus (immigrants tend to have slightly lower scores)
3. **GreekLanguageSkills** depends on both ImmigrantStatus and SelfEfficacy.

```{r simulate-data}
sim_data <- mottistefanidi2012[, c("IMM1", "SEF1", "GRK1")]
names(sim_data) <- c("ImmigrantStatus", "SelfEfficacy", "GreekLanguageSkills")
head(sim_data)
```

## Step 5: Examine Correlations Before and After Controlling for the Collider

We’ll demonstrate the difference between the **zero-order** correlation (i.e., unadjusted) between **ImmigrantStatus** and **SelfEfficacy** versus the partial correlation when controlling for **GreekLanguageSkills**.

```{r zero-order-corr}
# Zero-order correlation
cor_imm_se <- cor(sim_data$ImmigrantStatus, sim_data$SelfEfficacy)
cat("Zero-order correlation between Immigrant Status and Self-Efficacy:", round(cor_imm_se, 3))
```

```{r partial-corr}
# Partial correlation controlling for GreekLanguageSkills
# We'll compute the partial correlation by a linear model approach.
model_unadjusted <- lm(SelfEfficacy ~ ImmigrantStatus, data = sim_data)
model_adjusted   <- lm(SelfEfficacy ~ ImmigrantStatus + GreekLanguageSkills, data = sim_data)

summary(model_unadjusted)
summary(model_adjusted)
```

Compare the effect of **ImmigrantStatus** on **SelfEfficacy** before and after adjusting for **GreekLanguageSkills**. In the real-world data described by Asendorpf, adjusting for the collider drastically reduced the relationship to nearly zero.

## Step 6: Interpreting Results

Because **GreekLanguageSkills** is a collider in this scenario, conditioning on it can induce spurious associations or mask real ones. The observed difference in effect sizes (or correlation) for **ImmigrantStatus** on **SelfEfficacy** before and after adjusting highlights the bias that arises when we statistically control for a collider.

## Conclusion

This vignette illustrates how DAGs can help researchers:
1. Identify colliders and other key variables.
2. Understand when controlling for certain variables can bias results.
3. Guide more principled decisions about statistical models.

By embracing DAGs, psychologists and social scientists can align their analyses with explicit causal theories, thus reducing errors and advancing a clearer understanding of complex phenomena. As Asendorpf (2012) and Lee emphasize, **making causal assumptions explicit**—even in purely correlational settings—pushes the field toward more robust and defensible conclusions.

## References

```{r refs, echo=FALSE}
# Use a standard bibstyle or a set of references. For simplicity:
cat(
"Asendorpf, J. B. (2012). Causal interpretations of correlations. *European Journal of Personality, 26*(5), 463-465.\n",
"Foster, E. M. (2010). Causality and developmental psychology. *Developmental Psychology, 46*(6), 1454-1480.\n",
"Lee, *Citation details if known*.\n",
"Motti-Stefanidi, F., Asendorpf, J. B., & Masten, A. S. (in press). Immigrant youth adaptation in Greece. *Journal or Book reference here*.\n"
)
```


