---
title: "Specifying Augmented Directed Acyclic Graphs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Specifying Augmented Directed Acyclic Graphs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(theorytools)
```

In this vignette, we introduce a specification for **augmented** Directed Acyclic Graphs (aDAGs).
We will overload regular DAGs as specified using the **dagitty** package in R.
While `dagitty` provides standard facilities for declaring nodes, edges, exposures, and outcomes,
we augment the DAG with additional **metadata fields** to make it more conducive to theory specification.
These metadata fields do not interfere with regular use of the DAG in `dagitty`.

These metadata fields include:

- **tags** for nodes, like `exposure`, `outcome`, and `unobserved`. This metadata field is used by `dagitty` for identifying nodes of interest in causal inference.
- **`pos`** for nodes, which define the layout position in the X and Y dimension, e.g., `pos="0,1` is positioned at coordinates X = 0 and Y = 1. This metadata field is used by `dagitty`.
- **`label`** for nodes or edges: A descriptive label used for visualization and reporting. This is a new metadata field.
- **`distribution`** for nodes: The assumed distribution-generating function for each node. For exogenous nodes, this constitutes the distribution of the node itself; for endogenous nodes, this constitutes the residual distribution. This is a new metadata field.
- **`form`** for edges: A function specification (in a form interpretable by `as.formula()`) that can describe how a child node is calculated from its parents. This is a new metadata field.

Throughout the vignette, we will illustrate how to write an augmented DAG, how to parse and inspect it with `dagitty`, `theorytools`, and `tidySEM` (for plotting), and how to leverage these additional properties for further modeling or simulation tasks.

## A Basic DAG

The usual syntax for specifying a DAG in the `dagitty` R-package is something like:

```{r}
library(dagitty)
dagitty("dag {
  X -> Y
  Z -> X
  Z -> Y
}")
```

There are several tags that can be used in `dagitty`. Note that quotation marks used in tags **must be** double quotes `"`, so it makes sense to wrap the whole DAG syntax in single quotes `'`:

```{r}
library(dagitty)
dagitty('dag {
  X [exposure, pos="0,1"]
  Y [outcome, pos="1,1"]
  Z [unobserved, pos="1,0"]
  X -> Y
  Z -> X
  Z -> Y
}')
```

## Augmented DAGs (aDAGs)

In our **augmented** specification, we add additional properties as metadata fields.

Below, we detail each new property:

### `label` (Nodes/Edges)

- **Usage**: Provide a descriptive name for the node or edge.
- **Example**: `X [label="Study hours"]`

The label is used, for example, by `tidySEM` to label nodes and edges:

```{r}
library(tidySEM)
g <- dagitty('dag {
  X [label="Predictor", pos="0,1"]
  Y [label="Outcome", pos="1,1"]
  X -> Y [label="effect"]
}')
graph_sem(g)
```


## 2. `distribution` (Nodes)

- **Usage**: References a function that generates data for exogenous variables, or that describes the residual distribution for endogenous variables.
The function can reference the argument `n` to determine sample size, otherwise this is done implicitly.

- **Examples**:
  - `X [distribution="rnorm"]`: Node `X` is an exogenous variable drawn from a normal distribution with default arguments.
  - `Y [distribution="rnorm"]`: Node `Y` has residuals assumed to be normally distributed with default arguments.
  
```{r}
g <- dagitty('dag {
  X [distribution="rbinom(size = 2, prob = .5)"]
  Y [distribution="rnorm()"]
  X -> Y [form=".2*X"]
}')
```

## 3. `form` (Edges)

- **Usage**: Provides a formula-like specification for how a child node depends on its parent node(s). It should be something that `as.formula()` can parse.
- **Examples**:

* `X -> Y [form=".2*X"]` indicates that `Y` is a linear function of `.2` times `X`
* `X -> Y [form="X:Z"]` indicates that `Y` depends on an interaction between `X` and `Z`
* `X -> Y [form="X^2"]` indicates that `Y` depends on a quadratic function of `X`

```{r}
g <- dagitty('dag {
  X [distribution="rbinom(size = 2, prob = .5)"]
  Y [distribution="rnorm()"]
  X -> Y [form=".2*X"]
}')
```



# Example: Augmented DAG Specification

Below is a simple, hypothetical DAG showing how to combine these ideas. This DAG posits:
- `X`: Number of study hours, an `exposure`. Values are randomly sampled from 1-20 hours.
- `Z`: Stress level, an exogenous covariate, exponentially distributed (i.e., right-skewed, most people are not very stressed).
- `Y`: Exam performance an outcome depending on `X` and `Z`, with normally distributed residuals.s

```{r augmentedDAG, echo=TRUE, comment=""}
library(dagitty)

g <- dagitty('dag {
  X [exposure, label="Study Hours", distribution="sample.int(n = 20, size = n)"]
  Z [label="Stress Level", distribution="rexp()"]
  Y [outcome, label="Exam Performance", distribution="rnorm()"]
  X -> Y [label="direct effect", form="0.5+X"]
  Z -> Y [label="indirect effect", form="2*Z"]
}')
```


## Parsing DAG Properties

Augmented DAGs are interoperable with `dagitty`, but the `dagitty` package is not natively aware of the additional metadata fields used in `theorytools`, like `distribution` or `form`.
To access the augmented properties of aDAGs, the `theorytools` package uses `tidySEM`.
The purpose of the `tidySEM` package is to plot graphs (structural equation models and DAGs) as `ggplot` objects,
which can be further customized using regular `ggplot2` code.
It contains parsing functions to extract nodes and edges from a variety of objects, including `dagitty` graphs.
The functions `get_nodes()` and `get_edges()` parse the nodes and edges of aDAGs, respectively:

```{r nodeProperties, echo=TRUE}
get_nodes(g)
get_edges(g)
```

# Interpreting `distribution` and `form` in Simulation

A primary motivation for these augmented properties is **simulation**.
For example, you might simulate data by:

1. Generating `X` from `sample.int(n)`.
2. Generating `Z` from `rexp(n)`.
3. Generating `Y` using a formula that includes `X` and `Z` plus a residual from `rnorm(n)`.

Code to simulate data in line with these metadata can be automatically generated:

```{r simulateData, echo=TRUE, eval=TRUE}
cat(simulate_data(g, run = FALSE), sep = "\n")
```

You can use this script, for example, to generate synthetic data and build a reproducible analysis pipeline for a Preregistration-As-Code [@peikertReproducibleResearchTutorial2021; @vanlissaComplementingPreregisteredConfirmatory2022].

## Additional Notes

1. **Syntax Quoting**: In R, there are rules about using quotes within quotes. The `dagitty` package only recognizes double quotes (`" "`) inside graph specifications. This means you must **wrap the graph specification text in single quotes** (`' '`). Alternatively, you can escape every double quote inside the graph specification, which is not recommended because it is a hassle.
2. **Multiple Parents**: If a node has multiple parents, you can either specify multiple edges with their own `form` properties or a single edge with a combined formula. They are combined, and unique terms are retained.
3. **Order of Declaration**: `dagitty` does not mind the order in which nodes are declared, but you’ll need a topological order (no cycles) for valid DAG generation and simulation. 
4. **Integration with `dagitty` Functions**: The standard `dagitty` functions (e.g., `adjustmentSets()`) only look for recognized tags like `exposure` and `outcome`. They ignore custom properties like `distribution` and `form`, but these do not interfere with normal usage.

<!-- # Conclusion -->

<!-- By enriching your DAG specifications with the **`label`**, **`distribution`**, **`form`**, and relevant **exposure/outcome** tags, you can keep a single source of truth for both: -->
<!-- - Causal structure (as recognized by `dagitty`), and -->
<!-- - Data-generating or modeling assumptions (extra properties for simulation or advanced uses). -->

<!-- This approach reduces duplication and potential mismatches between your DAG-based causal assumptions and the code used to generate or analyze data. We hope this augmented specification pattern helps you maintain clarity and reproducibility in your causal modeling workflows. -->
