---
title: "Simulating Data from a FAIR Theory"
output: theorytools:::webex_vignette
vignette: >
  %\VignetteIndexEntry{Simulating Data from a FAIR Theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(theorytools)
```

```{r eval=knitr::is_html_output(), results='asis', echo=FALSE}
# This quizz is only shown in HTML output format
theorytools:::quizz(
"The answer to this question is true." = TRUE,
"This multiple choice question has three answers." =
c("Correct", "Incorrect", "Not sure"),
"Provide an exact floating point answer of 0.81" = 0.81,
"Or give some wiggle room" = c(0.81, .1),
"How many words in this sentence" = 6L
)
```

We start with the specification by Van Lissa and colleagues, conducted as part of Fiedler & Glöckner's many-theorists project:

```{r}
SDT1 <- read.csv("../dev/sdt1.txt")
knitr::kable(SDT1[, 1:3])
```


Next, we specify these verbal IF/THEN statements into causal links between specific constructs.
Note that the numbers associated with the statements are retained,
so that it is clear which causal links are derived from which IF/THEN statements:

```{r}
getwd()
SDT2 <- read.csv("../dev/sdt2.txt")
knitr::kable(SDT2[, 1:3])
```

This specification reveals redundancies in the original theory.
Certain points were repeated, once phrased in a positive way, and once phrased in a negative way.
When the statements are reduced to their implied causal connections, the redundancies become apparent.
Now, we can simply remove the redundancies as follows:

```{r}
# Drop statement numbers
SDT2 <- SDT2[, 2:3]
# Remove redundant statements
SDT2 <- SDT2[!duplicated(SDT2), ]
knitr::kable(SDT2)
```

We can subsequently translate this table to a DAG:

```{r}
SDT <- dagitty::dagitty(
  paste0("dag {",
         paste0(SDT2$from, " -> ", SDT2$to, collapse = "\n"),
         "}"))
```





